A Wait-free Queue with Polylogarithmic Step Complexity

### 2023-01-07 podc_paper/algorithm.tex
finished draft of code for GC
minor edits; changed tournament tree to ordering tree

\subsection{Overview}
We use a \emph{tournament tree} to agree on a total ordering of the operations performed on the queue.
The tree is a static binary tree of height $\ceil{\log_2 p}$ with one leaf 
assigned to each process. 
Each tree node  stores an array of \emph{blocks}, where each block represents a 
sequence of enqueues and a sequence of dequeues.
See Figure \ref{tournament} for an example.
In this section, we use an infinite array of blocks in each node.
Section \ref{reducing} describes how to replace the infinite array by a representation with bounded space.

@@

\begin{figure}[t]
\input{tournamentTree.pdf_t}
\caption{An example tournament tree with four processes. 
We show explicitly the enqueue sequence and dequeue sequence represented by each block in the \fld{blocks} arrays of the seven nodes.  The leftmost element of each \fld{blocks} array is a dummy block.
Arrows represent the indices stored in \eleft\ and \eright\ fields of blocks (as described in Section \ref{sec:fields}).
The fourth process's Deq\sub{6} is still propagating towards the root.
The linearization order for this tree is
Enq(a) Enq(e) Deq\sub{2} $\mid$ Enq(b) Deq\sub{4} Deq\sub{5} $\mid$ Enq(d) Enq(f) Enq(h) Deq\sub{1} $\mid$ Enq(c) Deq\sub{3} $\mid$ Enq(g), where vertical bars indicate boundaries of blocks in the root.\label{tournament}}
\end{figure}
\here{Make sure two figures appear on same page of final version}

\begin{figure}[t]
\input{implicit.pdf_t}
\caption{\label{implicit}The actual, implicit representation of the tree shown in Figure \ref{tournament}.
The leaf blocks simply show the \fld{element} field.
Internal blocks show the \fld{sum\sub{enq}} and \fld{sum\sub{deq}} fields,
and \eleft\ and \eright\ fields are shown using arrows as before.
Root blocks also have the additional \fld{size} field.
The \fld{super} field is not shown.}
\end{figure}

@@

Each tournament tree \node\ has an infinite array called \fld{blocks}.
To simplify the code, each \node\ has an empty \block\ $B_0$ in \fld{blocks}[0].
We define $E(B_0)$ and $D(B_0)$ to be the empty sequence.
Each \node's \fld{head} index  stores the position in the \fld{blocks} array to be used
for the next attempt to append a \block.

If a block contained an explicit representation of its sequences of enqueues and dequeues,
it would take $\Omega(p)$ time to construct a block, which would be too slow for task \ref{construct}.
Instead, the block stores an implicit representation of the sequences.
We now explain how we designed the fields for this implicit representation. 
Refer to Figure \ref{implicit} for an example showing how the tree in Figure \ref{tournament} is actually represented, and Figure \ref{object-fields} for the definitions of the fields of \blocks\ and \nodes.

@@

\renewcommand{\algorithmiccomment}[1]{\hfill\eqparbox{COMMENTSINGLE}{\com\ #1}}
\begin{figure}
\here{Does \var{leaf} have to be shared for helping when doing GC?}
\begin{algorithmic}[1]
\setcounter{ALG@line}{1}
\Statex \linecomment{Shared variable}
\begin{itemize}
\item \typ{Node} \var{root} \Comment{root of a binary tree of \tt{Node}s with one leaf for each process}
\end{itemize}

@@

\end{algorithmic}
\caption{Objects used in tournament tree data structure \label{object-fields}}
\end{figure}

@@

\Function{void}{Propagate}{\typ{Node} \var{v}}
    \State \linecomment{propagate blocks from \var{v}'s children to root}
    \If{\bf{not} \Call{Refresh}{\var{v}}} \label{firstRefresh}  \hfill \com\ double refresh
        \State \Call{Refresh}{\var{v}} \label{secondRefresh}
    \EndIf
    \If{\var{v} \bf{is not} \var{root}} \hfill \com\ recurse up tree
        \State \Call{Propagate}{\var{v}.\fld{parent}}
    \EndIf
\EndFunction{Propagate}

>>>>

\subsection{Overview}
Our \emph{\ordering\ tree} data structure is used to agree on a total ordering of the operations performed on the queue.
It is a static binary tree of height $\ceil{\log_2 p}$ with one leaf 
for each process. 
Each tree node  stores an array of \emph{blocks}, where each block represents a 
sequence of enqueues and a sequence of dequeues.
See Figure \ref{orderingtree} for an example.
In this section, we use an infinite array of blocks in each node.
Section \ref{reducing} describes how to replace the infinite array by a representation with bounded space.

@@

\begin{figure}[t]
\input{tournamentTree.pdf_t}
\caption{An example \ordering\ tree with four processes. 
We show explicitly the enqueue sequence and dequeue sequence represented by each block in the \fld{blocks} arrays of the seven nodes.  The leftmost element of each \fld{blocks} array is a dummy block.
Arrows represent the indices stored in \eleft\ and \eright\ fields of blocks (as described in Section \ref{sec:fields}).
The fourth process's Deq\sub{6} is still propagating towards the root.
The linearization order for this tree is
Enq(a) Enq(e) Deq\sub{2} $\mid$ Enq(b) Deq\sub{4} Deq\sub{5} $\mid$ Enq(d) Enq(f) Enq(h) Deq\sub{1} $\mid$ Enq(c) Deq\sub{3} $\mid$ Enq(g), where vertical bars indicate boundaries of blocks in the root.\label{orderingtree}}
\end{figure}
\here{Make sure two figures appear on same page of final version}

\begin{figure}[t]
\input{implicit.pdf_t}
\caption{\label{implicit}The actual, implicit representation of the tree shown in Figure \ref{orderingtree}.
The leaf blocks simply show the \fld{element} field.
Internal blocks show the \fld{sum\sub{enq}} and \fld{sum\sub{deq}} fields,
and \eleft\ and \eright\ fields are shown using arrows as before.
Root blocks also have the additional \fld{size} field.
The \fld{super} field is not shown.}
\end{figure}

@@

Each \node\ of the \ordering\ tree has an infinite array called \fld{blocks}.
To simplify the code, \fld{blocks}[0] is initialized with an empty \block\ $B_0$, 
where $E(B_0)$ and $D(B_0)$ are empty sequences.
Each \node's \fld{head} index  stores the position in the \fld{blocks} array to be used
for the next attempt to append a \block.

If a block contained an explicit representation of its sequences of enqueues and dequeues,
it would take $\Omega(p)$ time to construct a block, which would be too slow for task \ref{construct}.
Instead, the block stores an implicit representation of the sequences.
We now explain how we designed the fields for this implicit representation. 
Refer to Figure \ref{implicit} for an example showing how the tree in Figure \ref{orderingtree} is actually represented, and Figure \ref{object-fields} for the definitions of the fields of \blocks\ and \nodes.

@@

\renewcommand{\algorithmiccomment}[1]{\hfill\eqparbox{COMMENTSINGLE}{\com\ #1}}
\begin{figure}
\begin{algorithmic}[1]
\setcounter{ALG@line}{1}
\Statex \linecomment{Shared variable}
\begin{itemize}
\item \typ{Node} \var{root} \Comment{root of a binary tree of \tt{Node}s with one leaf for each process}
\end{itemize}

@@

\end{algorithmic}
\caption{Objects used in the \ordering\ tree data structure \label{object-fields}}
\end{figure}

@@

\Function{void}{Propagate}{\typ{Node} \var{v}}
    \State \linecomment{propagate blocks from \var{v}'s children to root}
    \If{\bf{not} \Call{Refresh}{\var{v}}} \label{firstRefresh}  \hfill \com\ double refresh
        \State \Call{Refresh}{\var{v}} \label{secondRefresh}
    \EndIf
    \If{$\var{v} \neq \var{root}$} \hfill \com\ recurse up tree
        \State \Call{Propagate}{\var{v}.\fld{parent}}
    \EndIf
\EndFunction{Propagate}

### 2023-01-07 podc_paper/boundedspace.tex
finished draft of code for GC
minor edits; changed tournament tree to ordering tree

Known lock-free search trees have step complexity that includes a term linear in $p$ \cite{EFHR14,Ko20}.  
However, we do not require all the standard search tree functions.
Instead of a standard insertion, we allow a \op{Refresh}'s insertion to fail if another
concurrent \op{Refresh} succeeds in inserting a node, just as the CAS on line \ref{cas}
can fail if a concurrent \op{Refresh} does a successful CAS.
Thus, we can use a particularly simple concurrent RBT implementation.
A sequential RBT can be made persistent using the classic node-copying technique of 
Driscoll et al.~\cite{DSST89}:  all nodes are immutable, and operations on the 
tree are performed by making a new copy of each node $v$ that must be modified, as well
as each node along the path from the root to $v$.
The tree reachable from the new copy of the root is the result of applying the tree operation.
This only adds a constant factor to the running time of any routine designed for a (sequential) RBT.
Once a process has performed an update to the RBT representing the blocks of a node 
$v$ in the tournament tree, 
it uses a CAS to swing $v$'s pointer from the previous RBT root to the new RBT root.
A search in the RBT can simply read the pointer to the RBT root and perform a standard
sequential search on it.
Bashari and Woelfel ~\cite{DBLP:conf/podc/BashariW21} used persistent red-black trees in a similar way to implement a snapshot data structure.

@@

To maintain good amortized time per operation, we periodically do a garbage collection (GC) phase.
If a process performing a \op{Refresh} on the root finds that $\var{root}.\head$ is a multiple
of $p^2$, it does GC.
To determine which blocks can be thrown away, we keep an array $\var{last}[1..p]$ where 
each process writes the index of the last block in
\var{root} containing an enqueue whose element it dequeued.
To perform GC, a process reads $\var{last}[1..p]$ and finds the maximum entry $m$.
Then, it helps complete every other process's pending dequeue 
by computing the dequeue's response and writing it in the block in the leaf that represents the dequeue.
Once this helping is complete, it follows from the FIFO property of the queue that elements enqueued 
in $\var{root}.\fld{blocks}[1..m-1]$ have all been dequeued, so GC can discard all blocks in the root's RBT 
with indices smaller than $m$.
Fortunately, there is an efficient \op{Split} operation that can remove
all nodes with indices less than $m$ from an RBT in logarithmic time \cite[Sec.~4.2]{Tar83}.
Then, GC can work its way down the tournament tree, using \op{Split} operations to discard all subblocks
of the discarded root blocks.

@@

Each process takes $\tilde{O}(p)$ steps during a GC phase.\footnote{We use $\tilde{O}$ notation
to omit polylogarithmic factors; see Appendix \ref{reducing-details} for precise bounds.}
Since all processes may run GC,
a total of $\tilde{O}(p^2)$ steps is spent on a GC phase.
There are at least $p^2$ operations between two GC phases,  so
GC contributes only $\tilde{O}(1)$ steps per operation in an amortized sense.

>>>>

Known lock-free search trees have step complexity that includes a term linear in $p$ \cite{EFHR14,Ko20}.  
However, we do not require all the standard search tree functions.
Instead of a standard insertion, we allow a \op{Refresh}'s insertion to fail if another
concurrent \op{Refresh} succeeds in inserting a node, just as the CAS on line \ref{cas}
can fail if a concurrent \op{Refresh} does a successful CAS.
Thus, we can use a particularly simple concurrent RBT implementation.
A sequential RBT can be made persistent using the classic node-copying technique of 
Driscoll et al.~\cite{DSST89}:  all nodes are immutable, and operations on the 
tree are performed by making a new copy of each node $v$ that must be modified, as well
as each node along the path from the root to $v$.
The tree reachable from the new copy of the root is the result of applying the tree operation.
This only adds a constant factor to the running time of any routine designed for a (sequential) RBT.
Once a process has performed an update to the RBT representing the blocks of a node 
$v$ in the \ordering\ tree, 
it uses a CAS to swing $v$'s pointer from the previous RBT root to the new RBT root.
A search in the RBT can simply read the pointer to the RBT root and perform a standard
sequential search on it.
Bashari and Woelfel ~\cite{DBLP:conf/podc/BashariW21} used persistent red-black trees in a similar way to implement a snapshot data structure.

@@

To maintain good amortized time per operation, we periodically do a garbage collection (GC) phase.
If a process performing a \op{Refresh} on the root finds that $\var{root}.\head$ is a multiple
of $p^2$, it does GC.
To determine which blocks can be thrown away, we keep an array $\var{last}[1..p]$ where 
each process writes the index of the last block in
\var{root} containing an enqueue whose element it dequeued.
To perform GC, a process reads $\var{last}[1..p]$ and finds the maximum entry $m$.
Then, it helps complete every other process's pending dequeue 
by computing the dequeue's response and writing it in the block in the leaf that represents the dequeue.
Once this helping is complete, it follows from the FIFO property of the queue that elements enqueued 
in $\var{root}.\fld{blocks}[1..m-1]$ have all been dequeued, so GC can discard all blocks in the root's RBT 
with indices smaller than $m$.
Fortunately, there is an efficient \op{Split} operation that can remove
all nodes with indices less than $m$ from an RBT in logarithmic time \cite[Sec.~4.2]{Tar83}.
Then, GC can work its way down the \ordering\ tree, using \op{Split} operations to discard all subblocks
of the discarded root blocks.

@@

Each process takes $\tilde{O}(p)$ steps during a GC phase.\footnote{We use $\tilde{O}$ notation
to omit polylogarithmic factors; see Appendix \ref{reducing-details} for precise bounds.}
Since all processes may run GC,
a total of $\tilde{O}(p^2)$ steps is spent on a GC phase.
There are at least $p^2$ operations between two GC phases,  so
GC contributes only $\tilde{O}(1)$ steps per operation in an amortized sense.
Doing GC less frequently would yield better constant factors in the amortized number of steps per operation but use more space.

### 2023-01-07 podc_paper/conclusion.tex
minor edits; changed tournament tree to ordering tree

Our focus in constructing our queue was on optimizing step complexity for worst-case executions.
However, our approach does have a higher cost in the best case (for example, when an operation
runs in isolation).
It may be possible to make our queue adaptive by having an operation capture a starting node
in the tournament tree (as in \cite{DBLP:conf/stoc/AfekDT95}) rather than having a statically assigned leaf as its starting point.
Another way to make our queue more practical might be to use it as the slow path in the
fast-path slow-path methodology of Kogan and Petrank \cite{10.1145/2370036.2145835} to
get a queue that has good performance in practice while also having good worst-case step complexity.

>>>>

Our focus in constructing our queue was on optimizing step complexity for worst-case executions.
However, our approach does have a higher cost in the best case (for example, when an operation
runs in isolation).
It may be possible to make our queue adaptive by having an operation capture a starting node
in the \ordering\ tree (as in \cite{DBLP:conf/stoc/AfekDT95}) rather than having a statically assigned leaf as its starting point.
Another way to make our queue more practical might be to use it as the slow path in the
fast-path slow-path methodology of Kogan and Petrank \cite{10.1145/2370036.2145835} to
get a queue that has good performance in practice while also having good worst-case step complexity.

### 2023-01-07 podc_paper/correctness.tex
minor edits; changed tournament tree to ordering tree

After proving some basic invariants in Section \ref{sec::basicProperties},
we show in Section \ref{sec::propagating} that performing a double refresh at each node
is sufficient to propagate an operation to the root.
In Section \ref{sec::helpingCorrect} we show that the \op{getEnqueue} and \op{indexDequeue}
routines correctly navigate through the tournament tree.
Finally, we show that the implementation is linearizable in Section \ref{sec::linearizability}.

@@

\Eric{I'm using immutable to mean that they are written only when the object is created.  The \fld{super} field does change from null to some value after the block is created, so it 
is written once, rather than 0 times.  In other words, we could have initialized
 all the other fields when the Block constructor is called.}
A \typ{Block} object's fields, except for \fld{super}, are immutable:  they are written only 
when the block is created at line \ref{enqNew} or \ref{deqNew} (for a leaf's block) or lines \ref{initNewBlock}--\ref{computeLength} (for an internal node's block).  
Moreover, only a \op{CAS} at line \ref{setSuper1} can modify the \fld{super} field 
(from \nl\ to a non-\nl\ value), so it remains fixed once a value is stored in it.
Similarly, only a \op{CAS} at line \ref{cas} can modify an element of a node's \fld{blocks} array 
(from \nl\ to a non-\nl\ value), so once a block is stored in a node, it remains there forever.
Only a \op{CAS} at line \ref{incrementHead} can update a node's \head\ field by incrementing it,
which implies the following.

@@

\begin{corollary}\label{lem::noDuplicates}
For  $i\neq j$, $v.\fld{blocks}[i]$ and $v.\fld{blocks}[j]$ cannot both contain the same operation.
\end{corollary}
\begin{proof}
The operations contained in a block $B$ are those that appear in subblocks of $B$ in the leaves of the tree.
Since each process puts each of its operations in only one block of its own leaf, an operation 
cannot be in two different leaf blocks. 
By \Cref{lem::subblocksDistinct}, $v.\fld{blocks}[i]$ and $v.\fld{blocks}[j]$ have no subblocks in common, so the claim follows.
\end{proof}

@@

Next, we prove the \fld{super} field of a block is within one of the true index of the block's superblock.
\here{An alternative would be modify the code
so that the correct index of the superblock is computed before storing it in $B.super$.
In Advance, we could check if $h>v.parent.blocks[h_p].end_{left/right}$ and if so write $h_p+1$ instead of $h_p$ in $v.blocks[h].super$.  Then, in IndexDequeue there would be no need to correct the value read from the super field.  If we do this, the following lemma would be modified to prove that the value is perfectly accurate.}

>>>>

After proving some basic invariants in Section \ref{sec::basicProperties},
we show in Section \ref{sec::propagating} that performing a double refresh at each node
is sufficient to propagate an operation to the root.
In Section \ref{sec::helpingCorrect} we show that the \op{getEnqueue} and \op{indexDequeue}
routines correctly navigate through the \ordering\ tree.
Finally, we show that the implementation is linearizable in Section \ref{sec::linearizability}.

@@

A \typ{Block} object's fields, except for \fld{super}, are immutable:  they are written only 
when the block is created at line \ref{enqNew} or \ref{deqNew} (for a leaf's block) or lines \ref{initNewBlock}--\ref{computeLength} (for an internal node's block).  
Moreover, only a \op{CAS} at line \ref{setSuper1} can modify the \fld{super} field 
(from \nl\ to a non-\nl\ value), so it remains fixed once a value is stored in it.
Similarly, only a \op{CAS} at line \ref{cas} can modify an element of a node's \fld{blocks} array 
(from \nl\ to a non-\nl\ value), so once a block is stored in a node, it remains there forever.
Only a \op{CAS} at line \ref{incrementHead} can update a node's \head\ field by incrementing it,
which implies the following.

@@

\here{Might be able to get rid of this corollary and just cite previous lemma instead to save space}
\begin{corollary}\label{lem::noDuplicates}
For  $i\neq j$, $v.\fld{blocks}[i]$ and $v.\fld{blocks}[j]$ cannot both contain the same operation.
\end{corollary}
\begin{proof}
The operations contained in a block $B$ are those that appear in subblocks of $B$ in the leaves of the tree.
Since each process puts each of its operations in only one block of its own leaf, an operation 
cannot be in two different leaf blocks. 
By \Cref{lem::subblocksDistinct}, $v.\fld{blocks}[i]$ and $v.\fld{blocks}[j]$ have no subblocks in common, so the claim follows.
\end{proof}

@@

Next, we prove the \fld{super} field of a block is within one of the true index of the block's superblock.

### 2023-01-07 podc_paper/introduction.tex
working on GC code
minor edits; changed tournament tree to ordering tree

Our queue uses a tournament tree where each process has its own leaf.\here{Is "tournament tree" the right word here?  Maybe call it an ordering tree?}
Each process propagates its operations along the path from its leaf up to the root.
Operations in the root are ordered, 
and this order is used to linearize the operations and compute their responses.
\here{Either here or in related work section, talk about previous usage of tournament tree and how ours differs from it}
Helping  ensures wait-freedom and avoids the CAS retry problem:  a process propagating its operation 
to a node $v$ helps propagate all operations from both of $v$'s children.
Explicitly storing all operations in the nodes would be too costly.
Instead, we use a novel implicit representation of sets
of operations that allows us to quickly merge two sets from the children of a node,
and quickly access any  operation in a~set.
\here{Maybe say a little more about techniques used in the implementation, if space permits}

>>>>

Our queue uses a binary tree, called the \emph{\ordering\ tree}, where each process has its own leaf.
Each process propagates its operations along the path from its leaf up to the root.
Operations in the root are ordered, 
and this order is used to linearize the operations and compute their responses.
\here{Either here or in related work section, talk about previous usage of ordering tree and how ours differs from it}
Helping  ensures wait-freedom and avoids the CAS retry problem:  a process propagating its operation 
to a node $v$ helps propagate all operations from both of $v$'s children.
Explicitly storing all operations in the nodes would be too costly.
Instead, we use a novel implicit representation of sets
of operations that allows us to quickly merge two sets from the children of a node,
and quickly access any  operation in a~set.
\here{Maybe say a little more about techniques used in the implementation, if space permits}

### 2023-01-07 podc_paper/related.tex
minor edits; changed tournament tree to ordering tree

\emph{Restricted Queues.}
David introduced the first sublinear-time queue
\cite{DBLP:conf/wdag/David04}, but it works only for a single enqueuer.
It uses fetch\&increment and swap  instructions and takes $O(1)$ steps per operation, but
uses unbounded memory.  Bounding the space increases the steps per operation to $\Omega(p)$.
Jayanti and Petrovic gave a wait-free poly-logarithmic
queue~\cite{DBLP:conf/fsttcs/JayantiP05}, but it works only for a single dequeuer. 
We adapt their idea of processes using
a tournament tree to agree on a linearization of
operations.

@@

\emph{Fetch\&Increment Objects.}
There are sublinear-time implementations of another object
with consensus number 2.
Two papers \cite{ERW12,10.1007/978-3-642-41527-2_20}
implemented fetch\&increment objects from LL/SC objects
using a polylogarithmic number of steps per operation.
Our approach is similar to that of Ellen, Ramachandran and Woelfel \cite{ERW12},
which uses a tournament tree to keep track of the fetch\&increments
that have been performed, as in the universal construction of \cite{DBLP:conf/stoc/AfekDT95}.
However, our construction requires more intricate
data structures to represent sets of operations, since a queue's state cannot be represented as succinctly
as the single-word state of a fetch\&increment object.

>>>>

\emph{Restricted Queues.}
David introduced the first sublinear-time queue
\cite{DBLP:conf/wdag/David04}, but it works only for a single enqueuer.
It uses fetch\&increment and swap  instructions and takes $O(1)$ steps per operation, but
uses unbounded memory.  Bounding the space increases the steps per operation to $\Omega(p)$.
Jayanti and Petrovic gave a wait-free poly-logarithmic
queue~\cite{DBLP:conf/fsttcs/JayantiP05}, but it works only for a single dequeuer. 
Our \ordering\ tree is similar to the tree structure the use to agree on a linearization ordering.

@@

\emph{Fetch\&Increment Objects.}
There are sublinear-time implementations of another object
with consensus number 2.
Two papers \cite{ERW12,10.1007/978-3-642-41527-2_20}
implemented fetch\&increment objects from LL/SC objects
using a polylogarithmic number of steps per operation.
Our approach is similar to that of Ellen, Ramachandran and Woelfel \cite{ERW12},
which also uses a tree structure to keep track of the fetch\&increments
that have been performed, as in the universal construction of \cite{DBLP:conf/stoc/AfekDT95}.
However, our construction requires more intricate
data structures to represent sets of operations, since a queue's state cannot be represented as succinctly
as the single-word state of a fetch\&increment object.
